---
title: "Expected kin counts by type of relative in a two-sex multi-state time-varying framework"
output: 
  html_document:
    toc: true
    toc_depth: 1
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In this vignette, we'll demonstrate how `Joe's code so far` can be used to compute parity-specific kinship networks for an average member of a given population, the sex of whom is user specified, subject to time-varying demographic rates. We call this individual Focal. 

We will compute the number, age, and parity distribution of Focal's relatives for each age of Focal's life, and as a function of the year in which Focal is born. 

## Kin counts by parity ##########################################################################

In this example we use data for the UK ranging from 1965 - 2005. The data can be obtained from the [Human Mortality Database](https://www.mortality.org/) and [Human Fertility Database](https://www.humanfertility.org/). 

Some simplifying assumptions we make due to current data availability are

i) Fertility rates vary with time, are distinct among parity class, but the same over sexes. I.e., we use the so-called ``androgynous approximation''.
ii) Mortality rates vary with time and are distinct across sex, but are the same across parity class (I have not implemented parity-specific mortality)
iii) The age-specific probabilities of parity-progression vary with time, but are the same over sex (androgynous approximation again)


In order to implement the model, the function `kin_multi_stage_TV_2_sex` expects the following 7 inputs, fed in as lists:

1) LIST: Female age-and-parity specific survival probabilities over the timescale. 
This input list has length = the timescale, and each entry represents the rates of a specific period in matrix form: stage columns, age rows.

2) LIST: Male age-and-parity specific survival probabilities over the timescale. 
This input list has length = the timescale, and each entry represents the rates of a specific period in matrix form: stage columns, age rows.

3) LIST: Female age-and-parity specific fertility rates over the timescale. 
This input list has length = the timescale, and each entry represents the rates of a specific period in matrix form: stage columns, age rows.

4) LIST: Male age-and-parity specific fertility rates over the timescale. 
This input list has length = the timescale, and each entry represents the rates of a specific period in matrix form: stage columns, age rows.

5) OUTER LIST with INNER LISTS: Female age-specific probabilities of moving up parity over the timescale. 
The outer list has length = the timescale. The inner list has length = number of ages. Each outer list entry is comprised of a list of matrices (stage*stage dimensional), each matrix describes age-specific probabilities of moving stage. Thus for each year, we have a list of age-specific probabilities of moving from one stage to the next.

6) Same as 5) but for males

7) LIST: Length = timescale, and each element is a matrix which assigns the offspring of individuals in some stage to the appropriate age class (age in rows and states in columns)

To avoid further calculations these lists are constructed in another file and simply imported below (See ``Examples'' , ``Matrix_construction'' for details)

```{r, message=FALSE, warning=FALSE}

# Lets construct these lists as model inputs..............
`%>%` <- magrittr::`%>%`


F_mat_fem <- readr::read_rds(here::here("Data","FF_list.Rds"))
F_mat_male <- readr::read_rds(here::here("Data","FM_list.Rds"))
T_mat_fem <- readr::read_rds(here::here("Data","T_list.Rds"))
U_mat_fem <- readr::read_rds(here::here("Data","UF_list.Rds"))
U_mat_male <- readr::read_rds(here::here("Data","UM_list.Rds"))
H_mat <- readr::read_rds(here::here("Data","H_list.Rds"))

```

# To recap, above lists of period-specific demographic rates are :

U_mat_fem: list of age by stage matrices, entries give female probability of survival. 
List starting 1965 ending 2022

U_mat_male: list of age by stage matrices, entries give female probability of survival. 
List starting 1965 ending 2022

F_mat_fem: list of age by stage matrices, entries give female fert, 
List starting 1965 ending 2022

F_mat_male == F_mat_fem: list of age by stage matrices, entries give male fert,
List starting 1965 ending 2022

T_mat_fem: list of lists of matrices: Each outer list entry is a list of matrices where each matrix gives age-specific probabilities a female moves up parity (inner list has length of number of age-classes). 
Outer list starting 1965 ending 2022

T_mat_fem: list of lists of matrices: Each outer list entry is a list of matrices where each matrix gives age-specific probabilities a male moves up parity (inner list has length of number of age-classes). 
Outer list starting 1965 ending 2022 (we don't have parity-specific fert for males)

H_mat: list of matrices which redistributes newborns to age-class 1 and parity 0. No time-variation.

Next we feed them into the matrix model starting from 1965 and up to 1985. Other arguments are:
 
UK sex ration --> 0.51
specific_kin = F --> we want all of Focal's network
We are considering parity --> parity = T
Focal is female --> sex_Focal = "Female"
Accumulated kin in this example --> dist_output = FALSE
Focal born into parity 0 --> stage_Focal = 1
age-increments by 1 year --> nc = 1
timescale from 1965-1985 -- > seq(1965, 1965 + 20)

### 1. For now lets simply consider the accumulated number of kin Focal expects over the lifecourse

```{r, message=FALSE, warning=FALSE}
# Run kinship model for a female Focal over a timescale of no_years where
no_years <- 20
# and we start projecting kin in 1965
# We decide here to count accumulated kin by age of Focal, and not distrubtions of kin
kin_out_1965_1985 <- 
  kin_multi_stage_TV_2_sex(U_mat_fem[1:no_years],
                           U_mat_male[1:no_years],
                           F_mat_fem[1:no_years],
                           F_mat_male[1:no_years],
                           T_mat_fem[1:no_years],
                           T_mat_fem[1:no_years],
                           H_mat[1:no_years],
                           alpha = 0.51, ## Sex ratio -- UK value 
                           specific_kin = FALSE,
                           parity = TRUE,
                           dist_output = FALSE,
                           sex_Focal = "Female", ##  define Focal's sex at birth 
                           stage_Focal = 1, ## Define Focal's stage at birth 
                           nc = 1,
                           seq(1965, (1965 + no_years)))
```

### 1.1. Visualizing kin ##################################

Let us now visualize the distribution of relatives over Focal's lifecourse using the model output

### 1.1.1. Plotting kin for an average Focal at some fixed period in time ###

For an example, let's say we really want to understand the age*parity distributions of the accumulated number of
aunts and uncles older than Focal's mother and father, for each age of Focal, and for each period in time from 1965-1985. Some people will do.

We can visualise using the below code

```{r, fig.height=6, fig.width=8}
kin_out_1965_1985 %>% 
  dplyr::filter(group == "oa", 
                year %in% c(1965, 1970, 1975, 1980, 1985)) %>%
  ggplot2::ggplot(ggplot2::aes(x = Age_Focal, y = pred_no_kin, color = Stage_Kin, fill = Stage_Kin)) +
  ggplot2::geom_bar(position = "stack", stat = "identity") + 
  ggplot2::facet_grid(Sex~year) +
  ggplot2::scale_x_continuous(breaks = c(0,10,20,30,40,50,60,70,80,90,100)) + 
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5)) + 
  ggplot2::ylab("Older aunts and uncles") 
```


### 1.1.2. Plotting the kin of Focal as a function of Focal's cohort of birth #####

Since we only ran the model for 20 years (between 1968-1985), there is very little scope to view kinship as cohort-specific. We can however compare cohorts for 20-year segments of Focal's life.
Below we consider daughters

```{r, fig.height = 6, fig.width = 8}
kin_out_1965_1985 %>% 
  dplyr::filter(group == "d", cohort %in% c(1900,1925,1950,1965) ) %>%
  ggplot2::ggplot(ggplot2::aes(x = Age_Focal, y = pred_no_kin, color = Stage_Kin, fill = Stage_Kin)) +
  ggplot2::geom_bar(position = "stack", stat = "identity") + 
  ggplot2::facet_grid(Sex~cohort)  + 
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5)) + 
  ggplot2::ylab("Daughters") 
```
For example, the LHS plot should be interpreted as follows:

if Focal born in 1900, between 1965-1985 he/she will be 65-85 years old; offspring will be around 30-45, have completed reproduction/parity progression, and experiencing mortality.

the RHS simply refelcts the fact that Focal will not start reproduction until around 15 years old.















### 2. Now lets consider the distributions of kin Focal expects over the lifecourse (the abov)
```{r, message=FALSE, warning=FALSE}
kin_out_1965_1985_full <- 
  kin_multi_stage_TV_2_sex(U_mat_fem[1:no_years],
                           U_mat_male[1:no_years],
                           F_mat_fem[1:no_years],
                           F_mat_male[1:no_years],
                           T_mat_fem[1:no_years],
                           T_mat_fem[1:no_years],
                           H_mat[1:no_years],
                           alpha = 0.51, ## Sex ratio -- UK value 
                           specific_kin = FALSE,
                           parity = TRUE,
                           dist_output = TRUE,
                           sex_Focal = "Female", ##  define Focal's sex at birth 
                           stage_Focal = 1, ## Define Focal's stage at birth 
                           nc = 1,
                           seq(1965, (1965 + no_years)))
```

### 2.1. Visualizing kin ##################################

Let us now visualize the distribution of relatives over Focal's lifecourse using the model output

### 2.1.1. Plotting kin distributions for an average Focal of fixed age, at some fixed period in time ###

Below I plot the expected age*stage distribution of an average Focal's younger siblings over the years 1965, 1970, 1975, 1980 and 1985, given Focal is of age 50

```{r, fig.height = 6, fig.width = 8}
kin_out_1965_1985_full %>% 
  dplyr::filter(group == "ys", 
                year %in% c(1965, 1970, 1975, 1980, 1985),
                Age_Focal == 50) %>%
  ggplot2::ggplot(ggplot2::aes(x = Age_Kin, y = pred_no_kin, color = Stage_Kin, fill = Stage_Kin)) +
  ggplot2::geom_bar(position = "stack", stat = "identity") + 
  ggplot2::facet_grid(Sex~year) +
  ggplot2::scale_x_continuous(breaks = c(0,10,20,30,40,50,60,70,80,90,100)) + 
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5)) + 
  ggplot2::ylab("Younger nieces and nephews") +
  ggplot2::ggtitle("Focal 50")
```

As expected, the discontinuity reflects the fact that Focal's younger siblings cannot be of age >=50
